---
title: "Linear Regression and Centrality"
output: html_notebook
---

## Setup

We are going to discuss a couple of different data sets and a couple of different ideas tonight.

```{r setup, include=FALSE}
library(modelr)
library(tidyverse)
options(scipen=999)
data_file <- "./data/new-york-city-airbnb-open-data.zip"
airbnb <- read_csv(data_file)
```

## AirBNB

Earlier in the semester we introduced the AirBNB data set.

```{r}
airbnb
```

What about the price? What is the average and median prices of AirBNBs?

```{r}
avg_price <- mean(airbnb$price)
med_price <- median(airbnb$price)
lower_95 <- avg_price - sd(airbnb$price) * 2
upper_95 <- avg_price + sd(airbnb$price) * 2

bind_rows(
airbnb %>%
  group_by(neighbourhood_group) %>%
  summarize(lower_95 = round(mean(price) - sd(price)*2,0),
            avg_price = round(mean(price),0),
            upper_95 = round(mean(price) + sd(price)*2,0),
            med_price = median(price)),
airbnb %>%
  summarize(neighbourhood_group = "New York City",
            lower_95 = round(mean(price) - sd(price)*2,0),
            avg_price = round(mean(price),0),
            upper_95 = round(mean(price) + sd(price)*2,0),
            med_price = median(price)) 
)
```

- You should notice right away that our 95% confidence interval includes costs below zero.
- If you can find me a BNB in Staten Island that will pay over $400 a night, I suggest you NOT take them up on the generous offer.
- So what does the data really look like?

```{r}
airbnb %>%
  ggplot(aes(x = price)) +
  geom_density() +
  geom_vline(aes(xintercept = avg_price), color = "red") +
  geom_vline(aes(xintercept = med_price), color = "green") +
  geom_vline(aes(xintercept = lower_95), color = "blue") +
  geom_vline(aes(xintercept = upper_95), color = "blue")
```

And, if we "zoom in" on price below $2,500 a night, things don't really look that much saner.

```{r}
airbnb %>%
  filter(price <= 2500) %>% 
  ggplot(aes(x = price)) +
  geom_density() +
  geom_vline(aes(xintercept = avg_price), color = "red") +
  geom_vline(aes(xintercept = med_price), color = "green") +
  geom_vline(aes(xintercept = lower_95), color = "blue") +
  geom_vline(aes(xintercept = upper_95), color = "blue")
```

- While less dramatically true, the mean and median are not well-aligned with the largest group of prices.
- And again, I don't think any single summary statistic can really explain the patterns we see in the data, like the bumps at 500 and 1000.
- This data is very multi-modal, which technically violates one of the key assumptions of the mean.
    - Ironically, highly multi-modal data are sometimes treated as continuous and we all just ignore the bumps.
    - Because, of course we do.

